---
title: "BUA 455 Lecture 14 - Intro to Model Selection"
author: "Penelope Pooler Eisenbies"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
---

### Upcoming Dates

- **HW 6 was due yesterday (3/6)** but grace period is extended until tonight at midnight.

- **HW 7 is due on Monday, 3/20**. 

   - Grace Period is extended until Wednesday 3/22 at midnight, because of Spring Break.
   
<br>
  
- **Quiz 2 is Thursday, March 30th**

- Thursday's Lecture (3/9) will include In-class Exercises using the Animals Data and your HW 7 data to help you make progress.

---

### Getting Started with Markdown

- Download Zipped R project 

- Open Zipped folder and copy internal folder (R Project) to a BUA 345 folder on your computer NOT IN DOWLOADS

- Click on .Rproj file to open project

- Click on `code_data_output` file to open it.

- Click on `BUA_345_Lecture_14.Rmd` to open it.

- Run Setup Chunk

---

### Setup

* The setup chunk shows the packages needed for this demo.   

* R will install specified packages if needed (only required once after R is installed)  

* R will load specified packaged (required every time you start a new R session)  

* The first time you run this code, R will install these packages which will be slow.  

* **If you get warnings, that's okay.**  

* If you get **error messages**, I (or TA), can help you.

---

### Setup Chunk for Lecture 14

```{r setup for Lecture 14, include = T}

# this line specifies options for default options for all R Chunks
knitr::opts_chunk$set(echo=T, highlight=T)
# suppress scientific notation
options(scipen=100)

# install helper package that loads and installs other packages, if needed
if (!require("pacman")) install.packages("pacman", repos = "http://lib.stat.cmu.edu/R/CRAN/")

# install and load required packages
pacman::p_load(pacman,tidyverse, magrittr, olsrr, gridExtra, ggiraphExtra, knitr)

# verify packages
p_loaded()

```

**NOTES:

- ** Don't worry about `xaringanthemer` package (required for my slides but not for your code).

- If you are having trouble installing/loading any packages, please come to office hour or make an appointment with me or course TA.

---

### Lecture 14 In-class Exercises - Review Question

#### **Question 1 (L14) - Session ID: bua345s23**

- Recall the Actors and Athletes data that we examined in Lecture 14.

```{r celeb_prof regression}

# import and examine celeb profession dataset
celeb_prof <- read_csv("celeb_prof.csv", show_col_types=F) 

# formatted regression output - saved and printed to screen
(celeb_interaction_ols<- ols_regress(Earnings ~ Age + Profession + Age*Profession,
                                     data=celeb_prof, iterm = T))

```

---

### Lecture 14 In-class Exercises - Review Question

#### **Question 1 (L14) - Session ID: bua345s23**

#### Abridged Output

Review Question: Based on the regression output for the Actors and Athletes data, **what is the slope for the linear model for Athletes**?

Round answer to two decimal places.

Hint: To answer this question, you combine two terms:
-   the baseline slope term for `Age`: 1.824
-   the difference in slope `Athlete`: -5.063
-   Slope for Athletes = baseline + difference = `____`

---

#### Example of R<sup>2</sup> Interpretation

#### Import and Examine Insurance Data

```{r import and examine insure_L14 data}

insure <- read_csv("insure_L14.csv", show_col_types=F) # import
insure <- insure |> # create log transformed variable
  mutate(ln_charges = log(Charges)) |> glimpse(width=75)

```

---

#### Examine Histograms of Charges and ln_charges

```{r creatingand formatting histograms of Charges and ln_charges, echo=F, fig.width=15, fig.height=8, fig.align='center'}

# histogram of original Charges data
hist_Charges <- insure |> 
  ggplot() +  
  geom_histogram(aes(x=Charges), color="darkblue", fill="lightblue") + 
  labs(x="Insurance Charges", y="Frequency") +  
  theme_classic() + 
  theme(axis.title = element_text(size=18),
        axis.text = element_text(size=15),
        plot.background = element_rect(colour = "darkgrey", fill=NA, size=2))

# histogram of ln_charges
hist_ln_charges <- insure |> 
  ggplot() +  
  geom_histogram(aes(x=ln_charges), color="darkgreen", fill="lightgreen") + 
  labs(x="Natural Log of Insurance Charges", y="Frequency") +  
  theme_classic() + 
  theme(axis.title = element_text(size=18),
        axis.text = element_text(size=15),
        plot.background = element_rect(colour = "darkgrey", fill=NA, size=2))

# display of these two histograms side by side
grid.arrange(hist_Charges, hist_ln_charges, ncol=2)

```

---

### SLR model - Predictor (X) Variable: Age

```{r slr model for insurance data}

# save and print slr model
(insure_slr <- ols_regress(ln_charges ~ Age, data=insure))

```


---

### More about R<sup>2</sup> and How It's Calculated

- R = 0.528 which indicates a moderate correlation between `Age` and `ln_charges` (Natural Log of Insurance Charges)

- R<sup>2</sup> = 0.279 which means that approximately 28% of the variability in `ln_charges` (Natural Log of Insurance Charges) is explained by `Age`.

- R<sup>2</sup> can also be calculated from the Sum of Squares output:

   - **SS<sub>TOT</sub>** (Total. Sum of Squares): **1130.474 (Total variability in Y)**
  
   - **SS<sub>REG</sub>** (Regression. Sum of Squares): **314.960 (Variability in Y explained by model)**
   
   - **SS<sub>RES</sub>** (Residual Sum of Squares): **815.514 (Variability in Y NOT explained by model)**

   - **R<sup>2</sup> = SS<sub>REG</sub> / SS<sub>TOT</sub> = 314.96/1130.474 = 0.279**
   
---

### Example MLR Model with Quantitative Interaction Term

-   One POSSIBLE model for these data (there are many):

```{r insurance model iwth quantitative interaction}

# save and print mlr model output
(insure_mlr_quant1 <- ols_regress(ln_charges ~ Age + BMI + Children + Age*Children, data=insure))
insure_model1 <- insure_mlr_quant1$model # save model parameters to use in calculations

```

---

### Interpreting and Evaluating Quantitative Interactions


- Two CORRECT Interpretation(s) of this interaction:

  1. The effect of age on insurance charges differs depending on how many children you have.

  2. The effect of number of children on insurance charges differs depending your age.
  
- Which interpretation the analyst emphasizes depends on the question being addressed.

---

### Lecture 14 In-class Exercises - Quantitative Interactions

#### **Question 2 (L14) - Session ID: bua345s23**

- Based on the R MLR output shown, is the interaction between Age and Number of Children useful in explaining differences in Insurance Charges?

---

### Lecture 14 In-class Exercises - Quantitative Interactions

#### **Question 3 (L14) - Session ID: bua345s23**

- Using this model, what is estimated insurance charge for 45 year old with a BMI of 26 and 2 children? Round to closest whole dollar. 

   - Calculation can be done in R or Excel. Here are values for Excel calculation:
   
      - `Age = 45`
      - `BMI = 26`
      - `Children = 2`
      - `Age*Children = 45*2 = 90`
   
---

#### **Question 3 (L14) - Session ID: bua345s23**

- Using this model, what is estimated insurance charge for 45 year old with a BMI of 26 and 2 children? Round to closest whole dollar. 

#### Using a Saved Model in R to find an Estimate

First specify values using **EXACT** variables names in model.
`new_obs` is a one row data set (tibble) with these values.

```{r create a dataset with 1 new observation}

Age <- 45      # specify values using variable names in model
BMI <- 26
Children <- 2
(new_obs <- tibble(Age, BMI, Children)) # new_obs is 1 row dataset

```

---

#### **Question 3 (L14) - Session ID: bua345s23**

- Using this model, what is estimated insurance charge for 45 year old with a BMI of 26 and 2 children? Round to closest whole dollar. 

#### Using a Saved Model in R to find an Estimate

Next, use `mutate` to create a new column in `new_obs`.  
  - `est_ln_charges` is the estimate created using the `lm` (linear model) command with the saved model `insure_model1` and the `predict` command

```{r use model to predict estimated ln_charges for new_obs}
              
(new_obs <- new_obs |>                  # create estimate_ln_charges
  mutate(est_ln_charges = lm(insure_model1) |> predict(new_obs)))

```


The last step is to back-transform the `est_ln_charges` value to create `est_charges`, the estimate of charges in dollars for this individual.

```{r back_transform est_ln_charges to dollars}

# (new_obs <- new_obs |> mutate(est_charges = ____(_____)))

```

---

### Lecture 14 In-class Exercises - Quantitative Interactions

#### **Question 4 (L14) - Session ID: bua345s23**

In the above model, all included terms appear to be useful to the model. Is the interaction between Age and BMI also useful in explaining the variability in insurance charges?

Examine the model output to answer this question.

```{r insure model with two quant interactions}

# save and print mlr model output
(insure_mlr_quant2 <- ols_regress(ln_charges ~ Age + BMI + Children + 
                                    Age*Children + Age*BMI, data=insure))

```

---

### Goodness of Fit - Adjusted R<sup>2</sup>

- Previous slides show two possible models for these data. ***There are 63 possible models*** with these X variables and all two way interactions.

- Today we will discuss **Adjusted R<sup>2</sup>** as one option to compare different models (We will cover other model comparison measures soon).
  
  - **Adjusted R<sup>2</sup>** adjusts R<sup>2</sup> DOWNWARD by adding a penalty for additional predictor variables.

    - R<sup>2</sup> (unadjusted) should NOT be used to compare MLR models.

    - Adding predictors will always increase R<sup>2</sup>, even if predictors are not useful.

    - Instead we adjust: We penalize model R<sup>2</sup> for each additional variable added.

    - Adjusted R<sup>2</sup> only increases if model fit improvement exceeds penalty for adding terms.

    - P-values for each term and change in Adjusted R<sup>2</sup> often agree (but not always)

    - As P, number of predictors increases, the penalty increases.

    - Adjusted R<sup>2</sup> = 1 - $\frac{(1-R^2)(n-1)}{n-P-1}$

**Students are not required to memorize this equation but you should understand what it is doing.**

---

### Examine All Possible Models using Adjusted R<sup>2</sup>

```{r code to examine all possible insurance models}

insure_full <- lm(ln_charges ~ Age + BMI + Children + Age*Children + # full model specified
                    Age*BMI + BMI*Children, data=insure)

insure_all_models <- ols_step_all_possible(insure_full)   # all possible models

insure_all_subset <- insure_all_models |>                 # specify model subset
  select(mindex, n, predictors, rsquare, adjr) |>         # determined based on           
  filter(mindex %in% c(1,5,6,8,10,22,42,44,47))           # understanding of data

(insure_subset_print1 <- insure_all_subset |>   # format table for printing and print
  select(n, predictors, rsquare, adjr) |>       # select useful columns
  mutate(rsquare = rsquare |> round(4),         # round values to 4 decimal places
         adjr = adjr |> round(4)) |>
  
  arrange(desc(adjr)) |>                        # reorder by adjust rsquared
  rename(`No. of Predictors`= n,                # rename table columns
         Predictors = predictors,
         `R<sup>2</sup>` = rsquare,
         `Adjusted R<sup>2</sup>` =  adjr) |>
  
  kable())

```

---

### Plan for This Week and HW 7

- In HW 7, you will examine the correlation matrix and then do simple versions of steps 3 and 6.
- In Lecture 14 (Today) and Lecture 15 (Thursday), we will look at a more complex example.

#### Animals Data - Question: What factors affect a mammal's sleep duration?

- Original dataset includes 56 distinct species, including man.

- Notes:

  - Two species of elephants were removed from the analysis.
  
    - Population was limited to animals under 1000 pounds.

  - Natural log (LN) transformed variables were added to original data.
  - Observations with missing values are removed below
  - Working dataset has 49 observations (49 different species)

```{r import data and remove missing values}

# import and examine data
animals <- read_csv("animals.csv", show_col_types=F) |>
  filter(!is.na(LifeSpan) & !is.na(Gestation)) 

```

---

### Glimpse of Animals Data

```{r data glimpse}

animals |> glimpse(width=75)

```

---

### Animals Data Dictionary - Description of Variables

```{r animals data dictionary table, echo=F}

Variable <- names(animals)

Type <- c("Nominal", rep("Quantitative", 8), rep("Ordinal", 3))

Description <- c("Name of Species",
                 "Total Sleep = sum of slow wave and paradoxical sleep (hrs/day)",
                 "Average Body Weight in kilograms",
                 "Natural Log of Body Weight",
                 "Average Brain Weight in grams",
                 "Natural Log of Brain Weight",
                 "Maximum Life Span in years",
                 "Natural Log of Life Span",
                 "Gestation Time in days",
                 "Predation Index (1 = least likely to be preyed upon, 5 = most likely)",
                 "Sleep Exposure Index (1 = least exposed while sleeping, 5 = most exposed",
                 "Overall Danger Index (1 = least danger from other animals, 5 = most most danger")

(animal_data_dictionary = tibble (Variable, Type, Description) |>
  kable())

```

#### Intuitvely, there is likely to be redundancy between `Predation`, `Exposure`, and `Danger`.

---

### Matrix of Scatterplots 

- Like correlation matrices, scatterplot matrices are a useful way to quickly examine all possible predictors.
- First let's look at untransformed quantitative variables.

  - Notice BodyWt, BrainWt, LifeSpan plots all show observations clustered in the corner.
  - That indicates that these dsitributions are skewed and should be log transformed.
  -   Could be verified with histograms

```{r matrix 1}

animal_mat1 <- animals |> select(TotalSleep, 
                                 BodyWt, 
                                 BrainWt, 
                                 LifeSpan, 
                                 Gestation)

pairs(animal_mat1)

```

---

### Scatterplot Matrix of Transformed Variables

- Next is same matrix with the LN transformed variables.

- Notice that distributions look better BUT some variables are highly correlated

  - Recall: If $R_{XY} \geq 0.8$ for two variables, they cannot both be in the model because they are **multicollinear**.

  - `LNBodyWt` and `LNBrainWt` should definitely both NOT be in the final model.

  - `LNBrainWt` and `LNLifeSpan` appear close to the cutoff and should not be in the final model together, if possible.
  
```{r matrix 2}

animal_mat2 <- animals |> select(TotalSleep, 
                                 LNBodyWt, 
                                 LNBrainWt, 
                                 LNLifeSpan, 
                                 Gestation)

pairs(animal_mat2)

animal_mat2 |> cor() |> round(2) |> kable()

```


The correlation matrix shows the numerical values of the correlations we see in the scatterplot.


---

### Scatterplot Matrix of Ordinal Variables

#### Lecture 14 In-class Exercises

#### **Question 4 (L14) - Session ID: bua345s23**

Which two ordinal categorical predictor variables appear to be multicollinear, i.e., highly correlated?

```{r matrix 3}

animal_mat3 <- animals |> select(TotalSleep, 
                                 Predation, 
                                 Exposure, 
                                 Danger)

pairs(animal_mat3)

animal_mat3 |> cor() |> round(2)

```

---

### Backward Elimination

**`1.`** Data examination and transformations completed

**`2.`** Create a full 'saturated' model with all potential predictor variables and interaction terms (This is subjective).

```{r animals full model}

 # convert ordinal variables to factors
animals <- animals |>       
  mutate(PredF = factor(Predation), 
         ExposF = factor(Exposure), 
         DangrF=factor(Danger))

# full model (subjective)
animals_full <- lm(TotalSleep ~ LNBodyWt + LNBrainWt + 
                     LNLifeSpan + Gestation + 
                     PredF + ExposF + DangrF + 
                     LNBodyWt*Gestation + LNLifeSpan*PredF + 
                     LNLifeSpan*ExposF + LNLifeSpan*DangrF, data=animals)

```

---

### Backward Elimination

**`3.`** Use 'Backward Elimination' to pare full model down to a preliminary model.

- We cast a ***wide net*** to start by specifying that erms will remain in model if p-value < 0.1.  

- Note that each category of each factor variable is shown making model look more complex than it is.

```{r animals model backward elim and output}

(animals_BE <- ols_step_backward_p(animals_full, prem = 0.1, progress = T))

```

---

### Backward Elimination - Next Steps

**`4.`** Examine predictors in preliminary model to confirm they are not too highly correlated with each other. 

- If correlation for two variables, $R_{XY} \geq 0.8$, then one variable should be excluded.

- Variables in preliminary model: : `LNBodyWt`, `LNLifeSpan`, `Gestation`, `PredF`, `DangrF`, `LNLifeSpan*PredF`

  - Recall that `PredF` (Predation) and `DangrF` (Danger) are highly correlated.
  - `PredF` is included in an interaction term so exclude `DangrF`.

**`5.`** If model was modified in Step 4, rerun model through Backward Elimination (not always needed).

**`6.`**  Interpret final model. 

- Adjusted R<sup>2</sup> = 0.655
- Model below looks complicated, but each animal is in only one Predation Category.
- Baseline Predation Category = 1


```{r final animal model}
# specify final model
(animals_final <- ols_regress(TotalSleep ~ LNBodyWt + LNLifeSpan + Gestation + 
                               PredF + LNLifeSpan*PredF, data=animals))
# save coefficients
animals_model <- animals_final$model

```

---


### Using Model to Find Estimates - Exporting Model and Data to Excel

- This model can be used to find model estimates and residuals for all animals.

- We will **ALSO** do these calculations are done in an Excel Spreadsheet to clarify each model component in estimate.

- Below we export the data for three species to examine how the model works

```{r exporting data to excel for demo}
animals_model_data <- animals |>             # create new dataset with model variables only
  select(Species, TotalSleep, LNBodyWt, LNLifeSpan, Gestation, PredF)

(three_species <- animals_model_data |>       # create mini dataset with three species
  filter(Species %in% c("Baboon", "Donkey", "ArcticFox")) |>
  write_csv("ThreeSpecies.csv"))

```

---

### Using Model to Find Estimates - Exporting Model and Data to Excel

- Model coefficients for calculations can be extracted and exported to Excel.

- Below We create a two column dataset listing each model component and it's beta coefficient.

- That dataset is exported as a .csv file for an in-class exercise.

```{r export model betas}

# examine and export model betas for worksheet
beta <- animals_final$betas
model_term <- names(beta)
(animal_betas <- tibble(model_term, beta) |> 
    write_csv("animal_betas.csv"))

```

#### Lecture 14 In-class Exercises
#### **Question 5 (L14) - Session ID: bua345s23**

What is the regression estimate of total sleep for 'Donkey'?

---

### Using Model to Find Estimates in R

- Model estimates can be calculated in R.

- The Excel Worksheet is used to demonstrate and clarify how those estimates are calculated.

- Students will be expected to calculate an estimate using a model with this level of complexity on Quiz 2.

```{r regression model estimates in R}
animals_model_data <- animals_model_data |>     # add model estimates to data
  mutate(Est_TotalSleep = lm(animals_model) |> 
           predict(animals_model_data) |> round(2))

animals_model_data <- animals_model_data |>          # calculate residuals
  mutate(Resid = TotalSleep - Est_TotalSleep) |>
  relocate(Est_TotalSleep, Resid, .after=TotalSleep) # reorder variables

head(animals_model_data, 4) |> kable()               # print first 4 rows

```

---


---

### **Key Points from Today**

- Regression modeling can be overwhelming 

  - Too many options
  
<br>

- Automating part of the variable selection process is helpful.

  - Should not be attempted in Excel.
  
<br>

- Result from Automated process is preliminary.
  
  - Analyst should tinker with these results.
  
<br>

- Backward Elimination is one effective method of model selection.

   - We will cover others.

<br>

You may submit an 'Engagement Question' about each lecture until midnight on the day of the lecture. **A minimum of four submissions are required during the semester.**

